Finally tagless, partially evaluated

### 2008-03-30 tagless/talk.tex
NJPLS talk

\newcommand{\authorstack}[3]{\begin{footnotesize}\begin{tabular}[t]{@{}c@{}}\begin{normalsize}#1\end{normalsize}\\[\jot]#2\\\texttt{#3}\end{tabular}\end{footnotesize}\ignorespaces}
\newsavebox\carette \sbox\carette{\authorstack{Jacques Carette}{McMaster University}{carette@mcmaster.ca}}
\newsavebox\oleg    \sbox\oleg   {\authorstack{Oleg Kiselyov}{FNMOC}{oleg@pobox.com}}
\newsavebox\ccshan  \sbox\ccshan {\authorstack{Chung-chieh Shan}{Rutgers University}{ccshan@rutgers.edu}}

@@

\author[Jacques Carette and Oleg Kiselyov and Chung-chieh Shan]{\usebox\carette \and \usebox\oleg \and \usebox\ccshan}
\date{APLAS\\30 November 2007}

@@

\begin{frame}{Simple type preservation}
    \def\bool{\only<7-8>{\ \alert{bool}}}
    \begin{tikzpicture}
        \draw node [lang] (lang)
          {\begin{overprint}[12em]
               \onslide<1>\makebox[12em]{\alert{Typed} source language}
               \onslide<2>\makebox[12em]{$3 \le 4$}
               \onslide<3-7>\makebox[12em]{\texttt{LEQ (LIT 3, LIT 4):\makebox[4em][l]{\bool\ term}}}
               \onslide<8>\makebox[12em]{\texttt{\alert{leq} (\alert{lit} 3, \alert{lit} 4):\makebox[4em][l]{\bool}}}
               \onslide<9>\makebox[12em]{Simply typed $\lambda$-calculus}
           \end{overprint}};
        \draw [->] (lang) --
            node [above,midway,anchor=east]
              {\only<1>{\alert{Typed} metalanguage\hspace*{2em}}%
               \only<2-3>{evaluate}%
               \only<9>{\begin{tabular}{@{}r@{\qquad}}
                        Haskell constructor instances\\
                        or ML functors
                        \end{tabular}}}
            node [above,midway,anchor=east,text width=2.5in,inner xsep=0pt]
                {\only<-3,9>{\color{bg}}\ttfamily
                \alt<8->{\only<8>{\alert<8>{lit:\ int -> int}\\
                lit i = i\\[1ex]
                \alert<8>{leq:\ int * int -> bool}\\
                leq (i,j) = i <= j}}
                {\only<3-7>{\alert<7>{eval:\ \only<7>{'a }term -> \only<7>{'a }value}\\
                eval (LIT i) = INT i\\
                eval (LEQ (e1,e2)) =\\
                \quad \alert<5>{match (eval e1, eval e2) with\\
                \quad (INT i, INT j) ->} BOOL (i <= j)}}}
            +(-100:4cm)
            node [anchor=north,text depth=0pt,text height=9pt]
              {\llap{\onslide<3-7>{\texttt{BOOL }}%
                     \onslide<3-8>{\texttt{true:}}%
                     \only<1>{\rlap{\alert{Typed} target languages}}%
                     \only<2>{\rlap{\enspace\textrm{true}}}%
                     \only<9>{\rlap{evaluate}}}%
               \makebox[3em][l]{\texttt{\bool\only<3-7>{\ value}}}};
        \draw [->] (lang) -- +(-80:3.5cm)
            node [anchor=north west,inner xsep=0pt] {\only<9>{\hspace*{-16pt}compile}};
        \draw [->] (lang) -- +(-65:3cm)
            node [anchor=north west,inner xsep=0pt] {\only<9>{specialize}};
        \draw [->] (lang) -- +(-50:2.5cm)
            node [anchor=north west,inner xsep=0pt] {\only<9>{\hspace*{-3pt}CPS transform}};
        \draw      (lang)    +(-35:2cm) node {\dots};
    \end{tikzpicture}

@@

\section{The object language}
\subsection{As a constructor class in Haskell}
\subsection{As a functor signature in ML}

\begin{frame}[t,fragile]{The object language\only<3-8>{ as a constructor class}\only<9-16>{ as a functor signature}}
\setbeamercovered{transparent=20}
\begin{overprint}
\onslide<1-2>
\medskip
\begin{proofrules}
    \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
    \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
    \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
    \[ \text{$n$ is an integer} \justifies n:\ZZ \]
    \uncover<1>{\[ \text{$b$ is a boolean} \justifies b:\BB \]}
    \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
    \uncover<1>{\[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]}
    \uncover<1>{\[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]}
    \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
\end{proofrules}
\onslide<3-8>
\begin{semiverbatim}
class Symantics \alert<3>{repr} where
     \alert<4,5>{int }\alert<5>{:\!: Int -> repr Int}
     \alert<4,6>{lam }\alert<6>{:\!: (repr a -> repr b) -> repr (a -> b)}
     \alert<4,6>{fix }\alert<6>{:\!: (repr a -> repr a) -> repr a}
     \alert<4,7>{app }\alert<7>{:\!: repr (a -> b) -> repr a -> repr b}
     \alert<4  >{add }:\!: repr Int -> repr Int -> repr Int
     \alert<4,8>{if_ }\alert<8>{:\!: repr Bool -> repr a -> repr a -> repr a}
\end{semiverbatim}
\onslide<9-16>
\begin{semiverbatim}
module type Symantics = sig type \alert<9>{\smash{('c,'a)}\,repr}
 val \alert<10,11>{int}\alert<11>{:\,int\,->\,('c,int)\,repr}
 val \alert<10,12>{lam}\alert<12>{:\,(('c,'a)\,repr\,->\,('c,'b)\,repr)\,->\,('c,'a->'b)\,repr}
 val \alert<10,12>{fix}\alert<12>{:\,('x\,->\,'x)\,->\,(('c,'a->'b)\,repr as 'x)}
 val \alert<10,13>{app}\alert<13>{:\,('c,'a\,->\,'b)\,repr\,->\,('c,'a)\,repr\,->\,('c,'b)\,repr}
 val \alert<10   >{add}:\,('c,int)\,repr\,->\,('c,int)\,repr\,->\,('c,int)\,repr
 val \alert<10,14>{if_}\alert<14>{:\,('c,bool)\,repr\,->\,(unit\,->\,'x)\,->\,(unit\,->\,'x)}
     \alert<10,14>{   }\alert<14>{ \,         \,    \,->\,(('c,'a)\,repr as 'x)}
end
\end{semiverbatim}
\end{overprint}

>>>>

\newcommand{\authorstack}[3]{\begin{footnotesize}\begin{tabular}[t]{@{}c@{}}\begin{normalsize}#1\end{normalsize}\\[\jot]#2\\\texttt{#3}\end{tabular}\end{footnotesize}\ignorespaces}
\newsavebox\carette \sbox\carette{\authorstack{Jacques Carette}{McMaster University}{carette@mcmaster.ca}}
\newsavebox\oleg    \sbox\oleg   {\authorstack{Oleg Kiselyov}{FNMOC}{oleg@pobox.com}}
\newsavebox\ccshan  \sbox\ccshan {\color{red}\authorstack{Chung-chieh Shan}{Rutgers University}{ccshan@rutgers.edu}}

@@

\author[Jacques Carette and Oleg Kiselyov and Chung-chieh Shan]{\usebox\carette \and \usebox\oleg \and \usebox\ccshan}
\date{NJPLS\\7 March 2008}

@@

\begin{frame}{Simple type preservation}
    \def\bool{\only<7-8>{\ \alert{bool}}}
    \begin{tikzpicture}
        \draw node [lang] (lang)
          {\begin{overprint}[12em]
               \onslide<1>\makebox[12em]{\alert{Typed} source language}
               \onslide<2>\makebox[12em]{$3 \le 4$}
               \onslide<3-7>\makebox[12em]{\texttt{LEQ (LIT 3, LIT 4):\makebox[4em][l]{\bool\ term}}}
               \onslide<8>\makebox[12em]{\texttt{\alert{leq} (\alert{lit} 3, \alert{lit} 4):\makebox[4em][l]{\bool}}}
               \onslide<9>\makebox[12em]{Simply typed $\lambda$-calculus}
           \end{overprint}};
        \draw [->] (lang) --
            node [above,midway,anchor=east]
              {\only<1>{\alert{Typed} metalanguage\hspace*{2em}}%
               \only<2-3>{evaluate}%
               \only<9>{\begin{tabular}{@{}r@{\qquad}}
                        Haskell constructor instances\\
                        or ML modules
                        \end{tabular}}}
            node [above,midway,anchor=east,text width=2.5in,inner xsep=0pt]
                {\only<-3,9>{\color{bg}}\ttfamily
                \alt<8->{\only<8>{\alert<8>{lit:\ int -> int}\\
                lit i = i\\[1ex]
                \alert<8>{leq:\ int * int -> bool}\\
                leq (i,j) = i <= j}}
                {\only<3-7>{\alert<7>{eval:\ \only<7>{'a }term -> \only<7>{'a }value}\\
                eval (LIT i) = INT i\\
                eval (LEQ (e1,e2)) =\\
                \quad \alert<5>{match (eval e1, eval e2) with\\
                \quad (INT i, INT j) ->} BOOL (i <= j)}}}
            +(-100:4cm)
            node [anchor=north,text depth=0pt,text height=9pt]
              {\llap{\onslide<3-7>{\texttt{BOOL }}%
                     \onslide<3-8>{\texttt{true:}}%
                     \only<1>{\rlap{\alert{Typed} target languages}}%
                     \only<2>{\rlap{\enspace\textrm{true}}}%
                     \only<9>{\rlap{evaluate}}}%
               \makebox[3em][l]{\texttt{\bool\only<3-7>{\ value}}}};
        \draw [->] (lang) -- +(-80:3.5cm)
            node [anchor=north west,inner xsep=0pt] {\only<9>{\hspace*{-16pt}compile}};
        \draw [->] (lang) -- +(-65:3cm)
            node [anchor=north west,inner xsep=0pt] {\only<9>{specialize}};
        \draw [->] (lang) -- +(-50:2.5cm)
            node [anchor=north west,inner xsep=0pt] {\only<9>{\hspace*{-3pt}CPS transform}};
        \draw      (lang)    +(-35:2cm) node {\dots};
    \end{tikzpicture}

@@

\section{The object language}
\subsection{As a constructor class in Haskell}
\subsection{As a module signature in ML}

\begin{frame}[t,fragile]{The object language\only<3-8>{ as a constructor class}\only<9-16>{ as a module signature}}
\setbeamercovered{transparent=20}
\begin{overprint}
\onslide<1-2>
\medskip
\begin{proofrules}
    \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
    \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
    \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
    \[ \text{$n$ is an integer} \justifies n:\ZZ \]
    \uncover<1>{\[ \text{$b$ is a boolean} \justifies b:\BB \]}
    \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
    \uncover<1>{\[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]}
    \uncover<1>{\[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]}
    \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
\end{proofrules}
\onslide<3-8>
\begin{semiverbatim}
class Symantics \alert<3>{repr} where
     \alert<4,5>{int }\alert<5>{:\!: Int -> repr Int}
     \alert<4,6>{lam }\alert<6>{:\!: (repr a -> repr b) -> repr (a -> b)}
     \alert<4,6>{fix }\alert<6>{:\!: (repr a -> repr a) -> repr a}
     \alert<4,7>{app }\alert<7>{:\!: repr (a -> b) -> repr a -> repr b}
     \alert<4  >{add }:\!: repr Int -> repr Int -> repr Int
     \alert<4,8>{if_ }\alert<8>{:\!: repr Bool -> repr a -> repr a -> repr a}
\end{semiverbatim}
\onslide<9-16>
\begin{semiverbatim}
module type Symantics = sig type \alert<9>{\smash{('c,'a)}\,repr}
 val \alert<10,11>{int}\alert<11>{:\,int\,->\,('c,int)\,repr}
 val \alert<10,12>{lam}\alert<12>{:\,(('c,'a)\,repr\,->\,('c,'b)\,repr)\,->\,('c,'a->'b)\,repr}
 val \alert<10,12>{fix}\alert<12>{:\,('x\,->\,'x)\,->\,(('c,'a->'b)\,repr as 'x)}
 val \alert<10,13>{app}\alert<13>{:\,('c,'a\,->\,'b)\,repr\,->\,('c,'a)\,repr\,->\,('c,'b)\,repr}
 val \alert<10   >{add}:\,('c,int)\,repr\,->\,('c,int)\,repr\,->\,('c,int)\,repr
 val \alert<10,14>{if_}\alert<14>{:\,('c,bool)\,repr\,->\,(unit\,->\,'x)\,->\,(unit\,->\,'x)}
     \alert<10,14>{   }\alert<14>{ \,         \,    \,->\,(('c,'a)\,repr as 'x)}
end
\end{semiverbatim}
\end{overprint}
